{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or, How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-03-01 thesis/landslide.tex
define dpor happens-before

% TODO: update this linkerino (repository name)
% TODO: add some backup linkeroos
The Bochs port of Landslide is likewise open-source and available at \url{https://github.com/bblum/bochs}.
% TODO: refresh
The HEAD commit at the time of writing is 8984021.
The Simics port uses Simics's proprietary API and is hence unlicensed and available upon request.

@@

@@

\subsection{Preemption point identification}
\label{sec:landslide-pps}
% TODO: talk about how w/in func and w/out func are processed
\subsection{State space estimation}
\label{sec:landslide-estimate}
\subsection{Dynamic Partial Order Reduction}
\label{sec:landslide-dpor}

@@

\begin{algorithm}[h]
	\ForEach{$I \in S$}{
		Execute($I$)
	}
\end{algorithm}

ICB's approach could likewise be summarized as follows:

\begin{algorithm}[h]
	\For{$n \in [0..\mathsf{max}_P(S)]$}{
		\ForEach{$I \in S, P(I) \le n$}{
			Execute($I$)
		}
	}
\end{algorithm}

@@

However, when combined with DPOR to prune equivalent interleavings,
DPOR's reduction might not be sound with respect to the subset of $S$ under $P \le n$.
Prior work introduced Bounded Partial Order Reduction (BPOR) \cite{bpor},
a compatibility extension to DPOR for ICB to address this problem.
I refer the reader interested in the algorithm's details to the citation;
this section will discuss Landslide's implementation.

% TODO: say how u do it

>>>>

% TODO: update this linkerino (repository name)
% TODO: add some backup linkeroos
The Bochs port of Landslide is likewise open-source and available at \url{https://github.com/bblum/bochs}.
% TODO: refresh
The HEAD commit at the time of writing is 8984021.
The Simics port uses Simics's proprietary API and is hence unlicensed and available upon request for educational use only.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\subsection{Preemption point identification}
\label{sec:landslide-pps}
% TODO: talk about how w/in func and w/out func are processed
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{State space estimation}
\label{sec:landslide-estimate}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Dynamic Partial Order Reduction}
\label{sec:landslide-dpor}

Landslide implements Dynamic Partial Order Reduction (DPOR) \cite{dpor}
to identify concurrent yet independent thread transitions
whose permutations can safely be pruned from the state space
while still testing all possible program behaviours.

The DPOR implementation consists of 3 parts:
computing memory conflicts, computing happens-before,
and tagging alternate branches to explore to drive the state space exploration.
The former two are computed as each preemption point is reached,
for the associated thread transition pairwise with all other preceding thread transitions.
The latter is computed at the end of each full interleaving executed, using the results of the two former,
and constitutes the bulk of the algorithm.

\subsubsection{Happens-before}

The happens-before relation expresses when two thread transitions can potentially be reordered,
or in other words, are concurrent (despite the serialized nature of the simulated execution).
This relation is expressed in the following definitions paraphrased from \cite{dpor}.

\begin{itemize}
	\item {\bf Enabled}: A transition $t$ is enabled in a state $s$ when a state $s'$ exists such that $s \xrightarrow{t} s'$ exists.
		In systems terms, the scheduler at state $s$ considers the thread corresponding to $t$ to be runnable.
	\item {\bf Dependent}: Two transitions $t_i$ and $t_j$ are dependent if
		\begin{itemize}
			\item $t_1$ is enabled in $s$ and $s \xrightarrow{t_1} s'$, and
			\item $t_2$ is enabled in $s$ but not enabled in $s'$, or vice versa.
		\end{itemize}
		In systems terms, either $t_1$ and $t_2$ belong to the same thread,
		or the execution of $t_1$ at $s$ causes $t_2$'s thread
		to change state from blocked to runnable or vice versa.
		Landslide computes this relation in {\tt enabled\_by()} in {\tt save.c}.
		\footnote{The original definition includes a second criterion that, from $s$,
		the executions $t_1 t_2$ and $t_2 t_1$ result in identical states;
		this captures the memory independence relation which I defer to the next subsection.}
	\item {\bf Happens-Before}: The happens-before relation for a transition sequence $S = t_1 \dots t_n$
		is the smallest relation $\rightarrow_S$ on ${1 \dots n}$ such that
		\begin{enumerate}
			\item if $i < j$ and $S_i$ and $S_j$ are dependent then $i \rightarrow_S j$, and
			\item $\rightarrow_S$ is transitively closed.
		\end{enumerate}
		Landslide computes this relation in the aptly-named {\tt compute\_happens\_before()} in {\tt save.c}.
\end{itemize}
The happens-before relation is a partial order expressing the scheduling constraints of a given interleaving.
All pairs of interleavings not included are subject to reordering,
and hence candidates for new thread interleavings to test.

Note that DPOR's notion of happens-before differs from the traditional distributed systems definition \cite{lamport-clocks}
or Pure Happens-Before;
rather, it coincides with condition 3 of Limited Happens-Before as defined in \sect{\ref{sec:background-datarace}}
(in fact, Landslide's Limited HB implementation simply reuses the same result computed for DPOR's purpose).

The memory conflict relation expresses when two thread transitions are dependent,
or in other words, when their behaviour could potentially vary depending which executes first.

When each preemption point is reached,
Landslide saves the current set of all memory accesses since the previous preemption point,
compares it to the accesses from all previous transitions by different threads
which weren't marked as happens-before in the previous section,
and then empties that set for the next thread transition ({\tt shimsham\_shm()} in {\tt save.c}).

The set intersection is implemented in {\tt mem\_shm\_interseect()} in {\tt memory.c}.
It checks for read/write and write/write pairs to the same address with an $O(\mathsf{max}(m,n))$ scan of both access sets (pre-sorted).
Whenever a conflict is identified, it also invokes the data race analysis (\sect{\ref{sec:landslide-datarace}}).
It checks for {\tt free()}/access conflicts as well as access pairs,
effectively treating deallocation of a heap block as a ``poisoning'' write to its entire contents,
which should conflict with accesses to any address therein.

\subsubsection{State space exploration}

\subsubsection{Example}

@@

\begin{algorithm}[h]
	\ForEach{$I \in S$}{
		Execute($I$)
	}
	\caption{Straightforward exploration ordering}
	\label{alg:not-icb}
\end{algorithm}

% TODO: make sure latex doesn't fuck up the algorithm placement
ICB's approach could likewise be summarized as follows:

\begin{algorithm}[h]
	\For{$n \in [0..\mathsf{max}_P(S)]$}{
		\ForEach{$I \in S, P(I) \le n$}{
			Execute($I$)
		}
	}
	\caption{ICB exploration ordering}
	\label{alg:icb}
\end{algorithm}

@@

First of all, note that Algorithm~\ref{alg:icb} is structured in a way that repeats interleavings
with fewer than $n$ preemptions that have already been checked in previous iterations of the outer loop.
This is because the number of preemptions in each branch is not known in advance;
rather, the state space must always be explored in an overall depth-first approach,
at best skipping too-preemptful interleavings as they are encountered.
As simple as it would be to state ``{\bf foreach} $I \in {\mathsf{sort}_P(S)}$'' in pseudocode,
implementing such an ordering would be much less straightforward. %
% TODO: post committee review
%\footnote{Translating conference papers' pseudocode into usable implementations is often quite difficult \cite{itg2}.}
The core ICB loop appears in {\tt time\_travel()} in {\tt landslide.c}.
%(although not explicitly structured as a C-style loop in the code)

Secondly, prior work \cite{bpor} has shown that
when combined with DPOR to prune equivalent interleavings,
DPOR's reduction might not be sound with respect to the subset of $S$ under $P \le n$.
That work introduced Bounded Partial Order Reduction (BPOR),
a compatibility extension to DPOR for ICB to address this problem.
To summarize, % TODO: summarize this once you write the DPOR section so you know what terminology to write in terms of
%I refer the reader interested in the algorithm's details to the citation;
%this section will discuss Landslide's implementation.
% TODO: say how u do it

### 2018-03-01 thesis/overview.tex
define dpor happens-before

\section{Data Race Analsyis}

@@

\subsubsection{P2}
The thread library project \cite{thrlib} has two main components: implementing concurrency primitives, and implementing thread lifecycle and management routines.
The required concurrency primitives are as follows:
\begin{itemize}
	\item Mutexes, with the interface {\tt mutex\_lock(mp)} and {\tt mutex\_unlock(mp)}, whose functionality is described earlier this chapter. Students may use any x86 atomic instruction(s) they desire, such as {\tt xchg}, {\tt xadd}, or {\tt cmpxchg}, and/or the {\tt deschedule}/ {\tt make\_runnable} system calls offered by the reference kernel.
	\item Condition variables, with the interface {\tt cond\_wait(cvp, mp)}, {\tt cond\_signal} {\tt (cvp)}, and {\tt cond\_broadcast(cvp)}. {\tt cond\_wait} blocks the invoking thread, ``simultaneously'' releasing a mutex which protects some associated state (atomically, with respect to other calls to signal or broadcast under that mutex).
		{\tt cond\_signal} and {\tt cond\_broadcast} wake one or all waiting threads.
		Students must use the {\tt deschedule} and {\tt make\_runnable} system calls to implement blocking (busy-waiting is forbidden), and typically include an internal mutex to protect the condition variable's state as well.
		The primary challenge of this exercise is ensuring the aforementioned atomicity between {\tt cond\_wait}'s unlock and deschedule, with respect to the rest of the interface.
	\item Semaphores, with the interface {\tt sem\_wait(sp)} and {\tt sem\_signal(sp)} (sometimes called {\em proberen} and {\em verhogen} in other literature). The semaphore can be initialized to any integer value; if initialized to 1, it behaves like a mutex.
		Students typically implement semaphores using mutexes and condition variables, not using atomic instructions or system calls directly.
	\item Reader-writer locks (rwlocks), with the interface {\tt rwlock\_lock(rwp, mode)} and {\tt rwlock\_unlock(rwp)}. {\tt mode} may be either {\tt RWLOCK\_READ} or {\tt RWLOCK\_}- {\tt WRITE}.
		Behaves as mutexes, but multiple readers may access the critical section simultaneously.
		Students typically implement rwlocks using mutexes and condition variables, not using atomic instructions or system calls directly.
\end{itemize}
The interface to each also includes an associated {\tt \_init()} and {\tt \_destory()} function.

@@

The Pintos kernel architecture \cite{pintos} is used at several universities, including Berkeley, Stanford, and the University of Chicago.
The Pintos basecode implements a rudimentary kernel, consisting of a context switcher, round-robin scheduler, locking primitives, and program loader.
upon which students add more features in several projects.
Most relevant to this thesis, the basecode provides the following functions/libraries, among others:
\begin{itemize}
	\item Semaphores (the basic concurrency primitive, implemented using direct scheduler calls): {\tt sema\_up}, {\tt sema\_down}, {\tt sema\_try\_down};
	\item Locks (which simply wrap a semaphore initialized to 1), {\tt lock\_acquire}, {\tt lock\_}- {\tt release}, {\tt lock\_try\_acquire};
	\item Condition variables (also implemented using scheduler calls): {\tt cond\_wait}, {\tt cond\_}- {\tt signal}, {\tt cond\_broadcast}, with the same semantics as Pebbles P2 condvars;
	\item Basic round-robin scheduling facilities: {\tt thread\_block} (a kernel-level analogue to Pebbles's {\tt deschedule}), {\tt thread\_yield}
	\item Kernel thread lifecycle management, {\tt thread\_create} and {\tt thread\_exit}, including stack space memory management;
	\item Interrupt and fault handlers;
	\item A page allocator, {\tt palloc\_get\_page}, {\tt palloc\_get\_multiple}, {\tt palloc\_}- {\tt free\_page}, {\tt palloc\_free\_multiple}
\end{itemize}
Both Pebbles and Pintos basecodes offer a standard C library including {\tt malloc}, string-formatting, printing, etc.

>>>>

\section{Data Race Analsyis}
\label{sec:background-datarace}

@@

\subsubsection{P2}
The thread library project \cite{thrlib} has two main components: implementing concurrency primitives, and implementing thread lifecycle and management routines.
The required concurrency primitives are as follows:
\begin{itemize}
	\item Mutexes, with the interface {\tt mutex\_lock(mp)} and {\tt mutex\_unlock(mp)}, whose functionality is described earlier this chapter. Students may use any x86 atomic instruction(s) they desire, such as {\tt xchg}, {\tt xadd}, or {\tt cmpxchg}, and/or the {\tt deschedule}/ {\tt make\_runnable} system calls offered by the reference kernel.
	\item Condition variables, with the interface {\tt cond\_wait(cvp, mp)}, {\tt cond\_signal} {\tt (cvp)}, and {\tt cond\_broadcast(cvp)}. {\tt cond\_wait} blocks the invoking thread, ``simultaneously'' releasing a mutex which protects some associated state (atomically, with respect to other calls to signal or broadcast under that mutex).
		{\tt cond\_signal} and {\tt cond\_broadcast} wake one or all waiting threads.
		Students must use the {\tt deschedule} and {\tt make\_runnable} system calls to implement blocking (busy-waiting is forbidden), and typically include an internal mutex to protect the condition variable's state as well.
		The primary challenge of this exercise is ensuring the aforementioned atomicity between {\tt cond\_wait}'s unlock and deschedule, with respect to the rest of the interface.
	\item Semaphores, with the interface {\tt sem\_wait(sp)} and {\tt sem\_signal(sp)} (sometimes called {\em proberen} and {\em verhogen} in other literature). The semaphore can be initialized to any integer value; if initialized to 1, it behaves like a mutex.
		Students typically implement semaphores using mutexes and condition variables, not using atomic instructions or system calls directly.
	\item Reader-writer locks (rwlocks), with the interface {\tt rwlock\_lock(rwp, mode)} and {\tt rwlock\_unlock(rwp)}. {\tt mode} may be either {\tt RWLOCK\_READ} or {\tt RWLOCK\_\allowbreak{}WRITE}.
		Behaves as mutexes, but multiple readers may access the critical section simultaneously.
		Students typically implement rwlocks using mutexes and condition variables, not using atomic instructions or system calls directly.
\end{itemize}
The interface to each also includes an associated {\tt \_init()} and {\tt \_destory()} function.

@@

The Pintos kernel architecture \cite{pintos} is used at several universities, including Berkeley, Stanford, and the University of Chicago.
The Pintos basecode implements a rudimentary kernel, consisting of a context switcher, round-robin scheduler, locking primitives, and program loader.
upon which students add more features in several projects.
Most relevant to this thesis, the basecode provides the following functions/libraries, among others:
\begin{itemize}
	\item Semaphores (the basic concurrency primitive, implemented using direct scheduler calls): {\tt sema\_up}, {\tt sema\_down}, {\tt sema\_try\_down};
	\item Locks (which wrap a semaphore initialized to 1), {\tt lock\_acquire}, {\tt lock\_\allowbreak{}release}, {\tt lock\_try\_acquire};
	\item Condition variables (also implemented using scheduler calls): {\tt cond\_wait}, {\tt cond\_\allowbreak{}signal}, {\tt cond\_broadcast}, with the same semantics as Pebbles P2 condvars;
	\item Basic round-robin scheduling facilities: {\tt thread\_block} (a kernel-level analogue to Pebbles's {\tt deschedule}), {\tt thread\_yield}
	\item Kernel thread lifecycle management, {\tt thread\_create} and {\tt thread\_exit}, including stack space memory management;
	\item Interrupt and fault handlers;
	\item A page allocator, {\tt palloc\_get\_page}, {\tt palloc\_get\_multiple}, {\tt palloc\_free\_page}, and {\tt palloc\_free\_multiple}
\end{itemize}
Both Pebbles and Pintos basecodes offer a standard C library including {\tt malloc}, string-formatting, printing, etc.
